---
title: "Power Analysis"
author: "Leslie"
date: "10/1/2021"
output: #slidy_presentation
  html_document:
    keep_md: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = FALSE,
	message = FALSE,
	warning = FALSE
)
#libraries
library(RODBC)
library(tidyverse)
library(knitr)
library(nortest) # provides anderson-darling test for normality
library(lmtest) # provides durbin-watson test for serial autocorrelation
library(emon) # various power test functions using simulation as well as an alternative mannkendall function
library(ggrepel) # for nice point labels
library(pwr) # for t test power
library(plotly)

source("functions.R")

# retrieve data
channel <- odbcDriverConnect('driver={SQL Server};server=anrgdb;database=WQdata;trusted_connection=true')
df_phos <- sqlQuery(channel, "SELECT LakeID, Year, Phos as Value, 'lmpPhos' as 'param' FROM dbo.LayData_AnnualMeans_Sta1 WHERE dayssampled_phos>=8")
df_secc <- sqlQuery(channel, "SELECT LakeID, Year, Secc as Value, 'lmpSecc' as 'param' FROM dbo.LayData_AnnualMeans_Sta1 WHERE dayssampled_secc>=8")
df_chla <- sqlQuery(channel, "SELECT LakeID, Year, Chla as Value, 'lmpChla' as 'param' FROM dbo.LayData_AnnualMeans_Sta1 WHERE dayssampled_chla>=8")
df_spr <- sqlQuery(channel, "SELECT LakeID, Year, SprTP as Value, 'SprTP' as 'param' FROM dbo.SprData_AnnualMeans_Sta1 WHERE SprTP is not null")
odbcClose(channel)

# calculate normality tests, linear model, etc.
# df <- rbind(df_phos, df_secc, df_chla, df_spr) %>%
#   mutate(LakeID=as.character(LakeID)) %>% mutate(param=as.character(param)) %>% mutate(YearC=Year-1980) %>%
#   nest_by(LakeID, param) %>%
#   mutate(nYears=list(nrow(data))) %>% filter(nYears>=5) %>%
#   mutate(mean=list(round(mean(data$Value),1))) %>% mutate(sd=list(round(sd(data$Value),1))) %>%
#   mutate(pvalue=list(round(pvalue(data$Value, data$YearC),4))) %>% mutate(cv=round(sd/mean,3)) %>%
#   mutate(cv_mean=round(sd/(sqrt(sd)*mean),3)) %>%
#   mutate(lm=list(lm(data=data, Value~YearC))) %>%
#   mutate(dw=list(dwtest(lm))) %>% mutate(dw.pvalue=as.numeric(dw[4])) %>%
#   mutate(normality=list(shapiro.test(data$Value))) %>% mutate(normality.pvalue=as.numeric(normality[2])) %>%
#   mutate(lognormality=list(shapiro.test(log(data$Value)))) %>% mutate(lognormality.pvalue=as.numeric(lognormality[2])) %>%
#   mutate(adnormality=(ifelse(nYears>7,list(ad.test(data$Value)), list(-1)))) %>% mutate(adnormality.pvalue=as.numeric(adnormality[2])) %>%
#   mutate(adlognormality=(ifelse(nYears>7,list(ad.test(log(data$Value))), list(-1)))) %>% mutate(adlognormality.pvalue=as.numeric(adlognormality[2])) %>%
#   mutate(cor.test.kendall=list(cor.test(data$Value, data$YearC, alternative="two.sided", exact=FALSE, method="kendall"))) %>%
#   mutate(kendall.pvalue=as.numeric(cor.test.kendall[3])) %>%
#   mutate(emon.mk=list(mannkendall(data$YearC, data$Value, nsims.mk=1000))) %>% mutate(emon.mk.pvalue=as.numeric(emon.mk[1]))
# save(df, file="data/df.Rdata")

# load data table with above calculations
load("data/df.Rdata")

#############
#COME BACK TO THIS FOR NORMALITY TESTS
#############

# filter data, last 10 years only
df.10yr <- rbind(df_phos, df_secc, df_chla, df_spr) %>% filter(Year >= 2011)


```

## Samples needed to achieve 80% power given annual mean and standard deviation

```{r}
# number of samples needed given annual mean and stdev to detect with 80% probability that a lake is greater than 12 ug/L
# note: pwr.t.test is from the pwr package
annual.mean <- seq(5,12,.5)
st.dev <- seq(1,10,1)
output <- data.frame(annual.mean=numeric(), st.dev=numeric(), effect.size=numeric(), n=numeric())
for (i in annual.mean) {
  for (j in st.dev) {
    d = (i-12)/j
    try <- tryCatch(
      {test <- pwr.t.test(sig.level=0.05, d=d, power=0.8, type="one.sample", alternative="less")}, error=function(e) {"error"}
    )
  suppressWarnings(
    if (try=="error") {
      n = 0
    } else {
      n = test$n
    }
  )
  output[nrow(output)+1,] <- c(i,j, d, n)
  }
}

ggplot(output, aes(x=annual.mean, y=n)) +
  geom_point(size=1.5, aes(color=as.factor(st.dev))) +
  geom_line(size=1, aes(color=as.factor(st.dev))) +
  xlab("annual.mean") +
  ylab("Number of Samples") +
  guides(color = guide_legend(reverse=TRUE, title="Standard\nDeviation" )) +
  #scale_y_continuous(breaks=seq(0,600,20)) +
  theme_bw() 
# get color






```


## What standard deviation should we use?

```{r}
# significance tests:  kendall using base r, kendall using emon (iterative), durbin-watson for serial correlation
df_sign <- df[c("LakeID","param", "nYears", "kendall.pvalue", "emon.mk.pvalue", "dw.pvalue", "normality.pvalue", "lognormality.pvalue", "adnormality.pvalue", "adlognormality.pvalue")] %>% mutate(kendall.sign=ifelse(kendall.pvalue<.05, "sign", "n.s.")) %>% mutate(emon.mk.sign=ifelse(emon.mk.pvalue<.05, "sign", "n.s.")) %>% mutate(dw.sign=ifelse(dw.pvalue<.05, "sign", "n.s.")) %>% mutate(normality.sign=ifelse(normality.pvalue<.05, "sign", "n.s.")) %>% mutate(lognormality.sign=ifelse(lognormality.pvalue<.05, "sign", "n.s.")) %>% mutate(adnormality.sign=ifelse(adnormality.pvalue<.05, "sign", "n.s.")) %>% mutate(adlognormality.sign=ifelse(adlognormality.pvalue<.05, "sign", "n.s."))

# counts of lmpPhos lakes with significant distribution tests
lmpPhosSign <- subset(df_sign[df_sign$param=="lmpPhos",])  # 61 lakes shapiro test
# nrow(subset(lmpPhosSign[lmpPhosSign$normality.sign=="sign",])) # 13 are not normal shapiro test
# nrow(subset(lmpPhosSign[lmpPhosSign$lognormality.sign=="sign",])) # 7 are not normal shapiro test
# nrow(subset(lmpPhosSign[lmpPhosSign$adnormality.sign=="sign",])) # 22 are not normal anderson-darling test
# nrow(subset(lmpPhosSign[lmpPhosSign$adlognormality.sign=="sign",])) # 19  are not normal anderson-darling test

```


### Standard deviation of annual means is correlated with annual mean phosphorus

```{r}
data <- df %>% data.frame() %>% unnest(mean) %>% unnest(sd) %>% select(LakeID, param, mean, sd) %>% filter(param=="lmpPhos") 

p <- ggplot(data, aes(x=sd, y=mean, label=LakeID)) +
  geom_point() +
  geom_smooth() +
  #geom_label_repel(aes(label=LakeID), max.overlaps = 40) +
  #geom_text(aes(label=LakeID)) +
  scale_y_log10(breaks=c(5,seq(10,50,10),100,300)) +
  ylab("Annual Mean TP ug/L") +
  xlab("Standard Deviation") +
  theme_bw()

ggplotly(p)

```

### Remove Shelburne and Arrowhead

```{r}
data <- df %>% data.frame() %>% unnest(mean) %>% unnest(sd) %>% select(LakeID, param, mean, sd) %>% filter(param=="lmpPhos") %>%
  filter(LakeID != "SHELBURNE" & LakeID != "ARROWHEAD MOUNTAIN")

p <- ggplot(data, aes(x=sd, y=mean, label=LakeID)) +
  geom_point() +
  geom_smooth(method="lm") +
  #geom_label_repel(aes(label=LakeID), max.overlaps = 40) +
  #geom_text(aes(label=LakeID)) +
  scale_y_log10(breaks=c(5,seq(10,50,10),100,300)) +
  ylab("Annual Mean TP ug/L") +
  xlab("Standard Deviation") +
  theme_bw()

ggplotly(p)

```

### What if we don't have anything but Spring TP - use Spring TP to establish standard deviation?

### Only a few lakes have annual mean datasets that are not normally distributed even after log transformation

```{r}
# note: data.frame() converts from tibble to data.frame to get rid of hidden grouping variable
tableData <-lmpPhosSign %>% data.frame() %>% select(LakeID, lognormality.sign) %>% filter(lognormality.sign=="sign") 
kable(tableData)
```
